Making Sequential Programs Deterministic

### 2019-03-25 design.tex
moving text around

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@OrderNonDet> indicates that
    a collection or an array will have the same elements in every execution, but in a
    possibly different order.  \<@OrderNonDet> may only be written on
    collections and arrays.
    A collection is any subtype of \<java.util.Collection>,
    \<java.util.Iterator>, or \<java.util.Map>, including user-defined types.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

@@

The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations
\begin{verbatim}
@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
\end{verbatim}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations 
(\<@OrderNonDet> is not allowed because it is invalid on primitive types). \<@PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type annotation \<@NonDet>.

>>>>

Subtyping rule:

$\infer{\Gamma \vdash x : \tau_2}{\Gamma \vdash x : \tau_1 & \tau_1 <: \tau_2}$

Typing rule for method calls:
\todo{typesetting}

$\forall i : i >= 0 \infer{method(\overline{a_i}) : \tau_r}{\tau_r \  method(\overline{\tau_i \  p_i}) & \overline{a_i : \tau_i}}$

\subsection{Determinism type hierarchy}\label{type-hierarchy}
The core of the determinism type system is the following type qualifiers (see Figure~\ref{fig-determinism-hierarchy}):
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<@NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<@Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}

In this type system, \<@Det> is a subtype of \<@NonDet> denoted as \codeid{@Det <: @NonDet}.

@@

The underlying checker-framework on which we build our determinism checker has support for
polymorphic annotations. When a user writes a polymorphic annotation on a method signature or a type parameter,
it indicates that it could resolve to any type in the type system depending on how it is used.
In the determinism checker, we define a polymorphic annotation \<@PolyDet>.
One of the most common locations for a polymorphic annotation is at a method signature.
Consider the following method declaration annotated with polymorphic annotations:
\begin{verbatim}
@PolyDet int foo(@PolyDet String param1, @PolyDet boolean param2) { }
\end{verbatim}
There are two valid instantiations for this declaration:
\begin{enumerate}
    \item @NonDet int foo(@NonDet String param1, @NonDet boolean param2) { }
    \item @Det int foo(@Det String param1, @Det boolean param2) { }
\end{enumerate}
This indicates that method \<foo> can be called with arguments having any of the \codeid{@Det or @NonDet} type annotations.
%   (\<@OrderNonDet> is not allowed because it is invalid on primitive types).
 \<@PolyDet> resolves to the least upper bound of
the actual types on arguments. For instance, if this method is called as \codeid{foo(@Det String arg1, @NonDet boolean arg2)}, the 
checker framework resolves the method declaration as \codeid{@NonDet int foo(@NonDet String param1, @NonDet boolean param2)}
causing the return type to have the type annotation \<@NonDet>.

\subsection{Handling order nondeterminism}
In addition to the type qualifiers described in~\ref{type-hierarchy}, the determinism checker 
defines the type qualifier \<@OrderNonDet>. This qualifier represents
a collection or an array that will have the same elements in every execution, but in a
possibly different order.  \<@OrderNonDet> may only be written on
collections and arrays.
A collection is any subtype of \<java.util.Collection>,
\<java.util.Iterator>, or \<java.util.Map>, including user-defined types. The relationship
of \<@OrderNonDet> with respect the the other two type qualifiers is \codeid{@Det <: @OrderNonDet <: @NonDet} as shown in Figure~\ref{fig-determinism-hierarchy}.
